---
title: "Speeding Stops & Arrests"
description: |
  Analysis of traffic stops data compiled by the Stanford Open Policing Project.
  
  Currently, a comprehensive, national repository detailing interactions between police and the public doesn’t exist. That’s why the Stanford Open Policing Project is collecting and standardizing data on vehicle and pedestrian stops from law enforcement departments across the country — and they’re making that information freely available. they’ve already gathered over 200 million records from dozens of state and local police departments across the country.

author: "Timeo Coletta"
date: October 5, 2025
format: 
  html:
    code-fold: true
    code-summary: "Show the code"
editor: visual
execute:
  warning: false
  message: false
---

## Introduction

In my analysis, I work with tables containing data on traffic stops made by state patrols in Rhode Island, North Carolina, and Massachusetts. Specifically, I examine whether there are differences across states in the proportion of arrests made during traffic stops with speeding as the stated reason for the stop. I consider the time period from 2007 to 2015, as the data tables overlap over this time frame. I hypothesize that Rhode Island has a higher proportion of arrests made for speeding stops than the two other states given its higher population density, which I infer makes for stricter speeding laws. I expect the difference to be smaller between Rhode Island and Massachusetts, given that the difference in their population density is very small.

```{r}
#load packages
library(tidyverse)
library(RMariaDB)
library(ggplot2)
library(gt)
```

```{r}
# set up SQL connection
con_traffic <- DBI::dbConnect(

  RMariaDB::MariaDB(),

  dbname = "traffic",

  host = Sys.getenv("TRAFFIC_HOST"),

  user = Sys.getenv("TRAFFIC_USER"),

  password = Sys.getenv("TRAFFIC_PWD")

)
```

```{sql connection = con_traffic, include=FALSE}
-- print names of tables in traffic database
SHOW TABLES;
```

```{sql connection = con_traffic, include=FALSE}
-- display unique names of rows in reason_for_stop column of rhode island table
SELECT DISTINCT reason_for_stop
FROM ri_statewide_2020_04_01
```

```{sql connection = con_traffic, include=FALSE}
-- print first 10 rows of rhode island table
SELECT * 
FROM ri_statewide_2020_04_01	 
LIMIT 0, 10;
```

```{sql connection = con_traffic, include=FALSE}
-- display unique names of rows in reason_for_stop column of north carolina table
SELECT DISTINCT reason_for_stop
FROM nc_statewide_2020_04_01;
```

```{sql connection = con_traffic, include=FALSE}
-- print first 10 rows of north carolina table
SELECT * 
FROM nc_statewide_2020_04_01 
LIMIT 0, 10;
```

```{sql connection = con_traffic, include=FALSE}
-- display unique names of rows in reason_for_stop column of massachussets table
SELECT DISTINCT reason_for_stop
FROM ma_statewide_2020_04_01;
```

```{sql connection = con_traffic, include=FALSE}
-- print first 10 rows of massachussets table
SELECT * 
FROM ma_statewide_2020_04_01 
LIMIT 0, 10;
```

## Data Wrangling

In the SQL query below, I calculate the yearly arrest percentages for traffic stops due to speeding in each state and join the results to create a data table with state-year pair observations. Storing the output table as an object in R allows me to use it later on for data visualization and statistical tests. Below is a preview of what the table looks like.

```{sql connection = con_traffic, output.var = "stops_table"}
SELECT
  'RI' AS state,
  YEAR(`date`) AS year,
  COUNT(*) AS total_stops,
  SUM(arrest_made) AS arrests,
  (SUM(arrest_made) / COUNT(*)) * 100 AS arrest_percentage
FROM ri_statewide_2020_04_01
WHERE reason_for_stop = 'Speeding' 
AND YEAR(`date`) BETWEEN 2007 AND 2015
GROUP BY YEAR(`date`)

UNION ALL

SELECT
  'NC' AS state,
  YEAR(`date`) AS year,
  COUNT(*) AS total_stops,
  SUM(arrest_made) AS arrests,
  (SUM(arrest_made) / COUNT(*)) * 100 AS arrest_percentage
FROM nc_statewide_2020_04_01
WHERE reason_for_stop = 'Speed Limit Violation'
AND YEAR(`date`) BETWEEN 2007 AND 2015
GROUP BY YEAR(`date`)

UNION ALL

SELECT
  'MA' AS state,
  YEAR(STR_TO_DATE(`date`, '%Y-%m-%d')) AS year,
  COUNT(*) AS total_stops,
  SUM(arrest_made) AS arrests,
  (SUM(arrest_made) / COUNT(*)) * 100 AS arrest_percentage
FROM ma_statewide_2020_04_01
WHERE reason_for_stop = 'Speed'
AND YEAR(`date`) BETWEEN 2007 AND 2015
GROUP BY YEAR(STR_TO_DATE(`date`, '%Y-%m-%d'))

ORDER BY state, year;
```

```{r}
#| fig-alt: "Table with columns state, year, total_stop, arrests, and arrest_percentage. 6 values for each column."
#| fig-cap: "Figure 1.1 - Stanford Open Policing Data"

# display first 6 rows of stops_table and print it in visually smooth format
head(stops_table) |> gt()

# terminate SQL connection
dbDisconnect(con_traffic)
```

## Plot #1 & Description

```{r}
#| fig-alt: "Plot with x axis year and y axis arrest percentage, with 3 lines each representing MA, NC, and RI. MA and RI show a downard trend over time while NC is relatively constant at 0.75."
#| fig-cap: "Figure 1.2 - Stanford Open Policing Data"

# create a vector matching each state to a color
state_colors <- c("RI" = "#E41A1C", "NC" = "#377EB8", "MA" = "#4DAF4A")

# create line plot comparing the proportion of arrests in each state over time
ggplot(stops_table, aes(x = year, y = arrest_percentage, color = state, group = state)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = state_colors) +
  labs(
    title = "Proportion of Arrests for Speeding Stops by State Over Time",
    x = "Year",
    y = "Arrest Percentage (%)",
    color = "State"
  ) +
  theme_minimal()

```

The above plot portrays the time series trend of arrest percentages for speeding-induced traffic stops in each respective state. It is interesting to note that North Carolina has higher consistency in arrest percentage over time than the other two states, who both trend downwards. This could signal that, in MA & RI, certain policies were enacted that reduced the chance of a speeding offense leading to arrest, such as more lenient regulations. North Carolina might have already had lenient regulations on speeding, so the proportion of arrests remains at a low point (whereas MA & RI could have been in transitional policy stages).

## Plot #2 & Description

**Comparison: RI vs MA**

-   Determining whether Rhode Island has a statistically significantly *higher* average arrest percentage for speeding stops compared to Massachusetts over the 2007-2015 period.

```{r}
#| fig-alt: "Two box plots, one representing MA and the other representing RI. with arrest percentage on the y axis. The p value is 0.3364, so the observed difference here is not statistically significant. RI centers around 1.25 while MA centers around 1.175."
#| fig-cap: "Figure 1.3 - Stanford Open Policing Data"

# isolates ri arrests
ri_arrests <- stops_table |>
  filter(state == "RI") |> pull(arrest_percentage)
   ma_arrests <- stops_table |>
     filter(state == "MA") |>
     pull(arrest_percentage)
   nc_arrests <- stops_table |>
     filter(state == "NC") |>
     pull(arrest_percentage)

# calculate p values based on t test
p_ri_ma <- t.test(ri_arrests, ma_arrests, alternative = "greater")$p.value
p_ri_nc <- t.test(ri_arrests, nc_arrests, alternative = "greater")$p.value

y_min <- 0.70
y_max <- 1.50

# create box plots comparing rhode island and massachussets
stops_table |>
  filter(state %in% c("RI", "MA")) |>
  ggplot(aes(x = state, y = arrest_percentage, fill = state)) +
  geom_boxplot(alpha = 0.7) +
  annotate("text", x = 1.5, y = y_max * 0.95, 
           label = paste0("p = ", round(p_ri_ma, 4)), 
           size = 5, fontface = "bold") +
  labs(
    title = "RI vs MA: Arrest Percentages for Speeding Stops (2007-2015)",
    x = "State",
    y = "Arrest Percentage (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("MA" = "#4DAF4A", "RI" = "#E41A1C")) +
  ylim(y_min, y_max)
```

The above plot showcases the difference in arrest percentages between Rhode Island and Massachusetts, with RI higher than MA. However, the p-value is greater than 0.05 which means the difference is not statistically significant (confidence interval of 95%), which I infer is because the population densities of both states are extremely similar (so I'd assume they have stricter regulations around speeding).

## Plot #3 & Description

**Comparison: RI vs NC**

-   Determining whether Rhode Island has a statistically significantly *higher* average arrest percentage for speeding stops compared to North Carolina over the 2007-2015 period

```{r}
#| fig-alt: "Two box plots, one representing NC and the other representing RI. with arrest percentage on the y axis. The p value is near 0, so the observed difference here is statistically significant. RI centers around 1.25 while MA centers around 0.75."
#| fig-cap: "Figure 1.4 - Stanford Open Policing Data"

# create box plots comparing rhode island and north carolina
stops_table |>
  filter(state %in% c("RI", "NC")) |>
  ggplot(aes(x = state, y = arrest_percentage, fill = state)) +
  geom_boxplot(alpha = 0.7) +
  annotate("text", x = 1.5, y = y_max * 0.95, 
           label = paste0("p = ", round(p_ri_nc, 4)), 
           size = 5, fontface = "bold") +
  labs(
    title = "RI vs NC: Arrest Percentages for Speeding Stops (2007-2015)",
    x = "State",
    y = "Arrest Percentage (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("NC" = "#377EB8", "RI" = "#E41A1C")) +
  ylim(y_min, y_max)
```

Based on the above plot, we can see that Rhode Island has a statistically significant (p \< 0.05) higher proportion of arrests for speeding-induced traffic stops than North Carolina. It is possible that North Carolina's much smaller population density (around 1/5) has prompted less serious penalties for speeding in the state, but I can only infer.

## Conclusion

In summary, the difference between RI and NC being statistically significant. Although analysis across a larger group of high density vs low density states would a beneficial area of exploration, it is difficult to conduct using this set of databases because state patrols have different ways of classifying reasons for stops. In fact, for some states (not included in this analysis) there is no clear "speeding" category, and the stated reason is simply moving violation (which encompasses a wide range of reasons). While I make inferences about speed limit regulations and population densities, these are purely guesses and differences in proportions could have a myriad of causal variables not measured here. It is interesting to note that so few arrests are made for speeding stops in every case, never exceeding 1.5% over the observed time period. This might be because speeding stops are nonviolent offenses and the vast majority are routine.

### **Sources**

<a href = "https://5harad.com/papers/100M-stops.pdf" target = "_blank" >Working Paper</a>

<a href = "https://openpolicing.stanford.edu/data/" target = "_blank" >Stanford Data</a>

#### Citation

Pierson, Emma, Camelia Simoiu, Jan Overgoor, Sam Corbett-Davies, Daniel Jenson, Amy Shoemaker, Vignesh Ramachandran, et al. 2020. “A Large-Scale Analysis of Racial Disparities in Police Stops Across the United States.” *Nature Human Behaviour*, 1–10
